### 2018-04-27 decision-tree.tex
updated
almost there

Building decision trees is a supervised learning process used to approximate a target function as a sequence of disjunctive conditions designed to measure the effectiveness of a set of attributes to classify data and predict outcomes. The theory behind decision trees is well established \citep[e.g.,][]{Quinlan_1986_ML, Mitchell_1997_Book}. 
\myrevision{A decision tree can be used to classify a case by starting at the root of the tree and moving through it until the leaf is encountered. Each tree is a structure including a root, decision nodes and leaf nodes. Decision nodes specifies some test to carried out on attributes and leaf nodes indicate the final class \citep{Quinlan_1993_Book}.} Decision trees are, however, non-unique. They depend on the dataset, the user parameters, and the algorithm employed to build them \citep{Murthy_1998_DMKD}. Therefore, before describing our choice of the decision tree algorithm and parameters, we revise three aspects about the dataset: data classes, attributes equivalence, and dataset balance.

@@

% SUGGESTED BY NAEEM:
% In the particular case of the C5.0 algorithm the result of this process depends on several parameters, logical and numerical. We set the options to do feature selection \myrevision{or winnowing. Therefore, the algorithm choose to use the most important attributes and ignore others. 
% We also activate the evaluation of possible advanced splits of data. Sometimes data is very close to the splitting threshold where assigning a hard threshold may incorrectly change the classification, whereas, that small difference could happen because of numerical errors or error in measurement. Therefore, advanced split can define a soft threshold to consider different probabilities in assigning data to different classes. 
% A decision tree can grow so deep to finally classify all training data correctly. However, the tree will be very complex and will suffer from overfitting. In other words it will have very poor functionality for test data. Therefore, decision tree algorithms (here C5.0) first grow the tree up to the most complex tree then based on user defined factors start to prune the grown tree. The pruning process helps to reduce the complexity of the tree and also increase it's generalizability. Consequently the pruned tree's leaf nodes will have training data which belongs to more than one class which the final class will be the mode of that leaf node. In this study we use two parameters to control the pruning process including confidence or certainty factor (CF) and the smallest number of samples that must be put in at least two of the splits ($S_{\min}$, also known as \textit{minCases}). C5.0 algorithm uses error-based pruning method (see \citet{Quinlan_1993_Book} for more details). The objective is to minimize the observed error rate. Therefore, error estimates for leaf nodes and subtrees are computed assuming that they were used to classify a set of unseen cases with a predicted error rate which is computed based on CF and binomial distribution of number of misclassified events (E) and all training data at that leaf (N). Decreasing CF and increasing $S_{\min}$ cause more heavy pruning process \citep[see][]{Kuhn_2017_Manual}.} In particular, we build a suite of trees for varying values of CF and $S_{\min}$.

>>>>

Building decision trees is a supervised learning process used to approximate a target function as a sequence of disjunctive conditions designed to measure the effectiveness of a set of attributes to classify data and predict outcomes. The theory behind decision trees is well established \citep[e.g.,][]{Quinlan_1986_ML, Quinlan_1993_Book, Mitchell_1997_Book}. 
% \myrevision{A decision tree can be used to classify a case by starting at the root of the tree and moving through it until the leaf is encountered. Each tree is a structure including a root, decision nodes and leaf nodes. Decision nodes specifies some test to carried out on attributes and leaf nodes indicate the final class \citep{Quinlan_1993_Book}.} 
Decision trees are, however, non-unique. They depend on the dataset, the user parameters, and the algorithm employed to build them \citep{Murthy_1998_DMKD}. Therefore, before describing our choice of the decision tree algorithm and parameters, we revise three aspects about the dataset: data classes, attributes equivalence, and dataset balance.

@@

% SUGGESTED BY NAEEM:
% In the particular case of the C5.0 algorithm the result of this process depends on several parameters, logical and numerical. We set the options to do feature selection \myrevision{or winnowing. Therefore, the algorithm choose to use the most important attributes and ignore others. We also activate the evaluation of possible advanced splits of data. Sometimes data is very close to the splitting threshold where assigning a hard threshold may incorrectly change the classification, whereas, that small difference could happen because of numerical errors or error in measurement. Therefore, advanced split can define a soft threshold to consider different probabilities in assigning data to different classes. A decision tree can grow so deep to finally classify all training data correctly. However, the tree will be very complex and will suffer from overfitting. In other words it will have very poor functionality for test data. Therefore, decision tree algorithms (here C5.0) first grow the tree up to the most complex tree then based on user defined factors start to prune the grown tree. The pruning process helps to reduce the complexity of the tree and also increase it's generalizability. Consequently the pruned tree's leaf nodes will have training data which belongs to more than one class which the final class will be the mode of that leaf node. In this study we use two parameters to control the pruning process including confidence or certainty factor (CF) and the smallest number of samples that must be put in at least two of the splits ($S_{\min}$, also known as \textit{minCases}). C5.0 algorithm uses error-based pruning method (see \citet{Quinlan_1993_Book} for more details). The objective is to minimize the observed error rate. Therefore, error estimates for leaf nodes and subtrees are computed assuming that they were used to classify a set of unseen cases with a predicted error rate which is computed based on CF and binomial distribution of number of misclassified events (E) and all training data at that leaf (N). Decreasing CF and increasing $S_{\min}$ cause more heavy pruning process \citep[see][]{Kuhn_2017_Manual}.} In particular, we build a suite of trees for varying values of CF and $S_{\min}$.

### 2018-04-27 introduction.tex
almost there

To that end, we study the relationships that exist between eleven different metrics---those proposed by \citet{Anderson_2004_Proc}, plus the strong motion duration---with the objective of offering a prioritized reduced number of metrics that can help predict validation results. We treat these metrics as independent variables defining a multi-dimensional space, and analyze them using machine learning techniques. We use a constrained \kmeans{} method \citep[e.g.,][]{Macqueen_1967_Proc, Wagstaff_2001_Proc} and conduct a subspace clustering analysis to address high-dimensional effects in order to add labels to the dataset in terms of four categories: poor, fair, good, and excellent. These categories \change{were defined by \citet{Anderson_2004_Proc} and have been adopted---despite some differences---by other GOF methods used in verification and validation \citep[e.g.,][]{Kristekova_2009_GJI, Olsen_2010_SRL}.}
% \myrevision{are commonly used in previous validation studies} 
% \myremove{adhere to the common practice in validation}. 
\myrevision{In the machine learning linguistic, \kmeans{} clustering approach is considered as an unsupervised learning method. However, when we add background knowledge into the process it becomes constrained \kmeans{} method and is considered as a semi-supervised learning method.} 
The dataset used corresponds to a series of simulations for the 2008 \eqmag{w} 5.4 Chino Hills, California, earthquake, and their corresponding validation results \citep{Taborda_2014_BSSA}. Upon labeling the dataset, we develop a family of decision trees using a C5.0 algorithm \citep[][]{Quinlan_1993_Book, Quinlan_1996_JAIR}, and select a few to prioritize the metrics involved in the validation process. The decision trees narrow the number of metrics offering a prediction algorithm into the aforementioned validation categories. \myrevision{Decision trees are considered as supervised learning algorithms.}

>>>>

To that end, we study the relationships that exist between eleven different metrics---those proposed by \citet{Anderson_2004_Proc}, plus the strong motion duration---with the objective of offering a prioritized reduced number of metrics that can help predict validation results. We \change{initially} treat these metrics as independent variables defining a multi-dimensional space, and analyze them using machine learning techniques. \change{In particular, we used semi-supervised and supervised methods. In machine learning, whether a method is considered unsupervised, semi-supervised or supervised, depends on the level of constraints applied to the process or prior level of classification given to the data. Here, we first use a semi-supervised method called \textit{constrained k-means} in order to conduct a subspace clustering analysis to label data samples in a given validation dataset \citep[e.g.,][]{Macqueen_1967_Proc, Wagstaff_2001_Proc}. This dataset corresponds to the validation results from a series of simulations done for the 2008 \eqmag{w} 5.4 Chino Hills, California, earthquake \citep{Taborda_2014_BSSA}. Here, the clustering done by the constrained $k$-means method allows us to label the data using for validation categories: poor, fair, good, and excellent. These categories were defined by \citet{Anderson_2004_Proc} and have been adopted---despite some differences---by other GOF methods used in verification and validation \citep[e.g.,][]{Kristekova_2009_GJI, Olsen_2010_SRL}. Upon labeling the data samples in our dataset into these four categories, we use a supervised learning method as implemented in the C5.0 algorithm to obtain a family of decision trees \citep[][]{Quinlan_1993_Book, Quinlan_1996_JAIR}. We then select a few of these trees to gain insight about and} prioritize the metrics involved in the validation process. The decision trees narrow the number of metrics offering a prediction algorithm into the aforementioned validation categories.

% We use a constrained \kmeans{} method \citep[e.g.,][]{Macqueen_1967_Proc, Wagstaff_2001_Proc} and conduct a subspace clustering analysis to address high-dimensional effects in order to add labels to the dataset in terms of four categories: poor, fair, good, and excellent. These categories \change{were defined by \citet{Anderson_2004_Proc} and have been adopted---despite some differences---by other GOF methods used in verification and validation \citep[e.g.,][]{Kristekova_2009_GJI, Olsen_2010_SRL}.} \myrevision{are commonly used in previous validation studies} \myremove{adhere to the common practice in validation}. \myrevision{In the machine learning linguistic, \kmeans{} clustering approach is considered as an unsupervised learning method. However, when we add background knowledge into the process it becomes constrained \kmeans{} method and is considered as a semi-supervised learning method.} The dataset used corresponds to a series of simulations for the 2008 \eqmag{w} 5.4 Chino Hills, California, earthquake, and their corresponding validation results \citep{Taborda_2014_BSSA}. Upon labeling the dataset, we develop a family of decision trees using a C5.0 algorithm \citep[][]{Quinlan_1993_Book, Quinlan_1996_JAIR}, and select a few to prioritize the metrics involved in the validation process. The decision trees narrow the number of metrics offering a prediction algorithm into the aforementioned validation categories. \myrevision{Decision trees are considered as supervised learning algorithms.}
